

_**Exercise:**_  
Run regressions for the other 3 studies by extending the code. Summarize the results for yourself like we did above.

```{r regression_ex, exercise=TRUE}

NSSpost_2 <- lm()
summary(NSSpost_2)

NSSpost_3 <- lm()
summary(NSSpost_3)

NSSpost_4 <- lm()
summary(NSSpost_4)

```

```{r regression_ex-solution}

NSSpost_2 <- lm(NSS7d.post ~ treatment, 
                  data = DatIAD[DatIAD$Study == "Hav19", ])
summary(NSSpost_2)

NSSpost_3 <- lm(NSS7d.post ~ treatment, 
                  data = DatIAD[DatIAD$Study == "Hav20AGED", ])
summary(NSSpost_3)

NSSpost_4 <- lm(NSS7d.post ~ treatment, 
                  data = DatIAD[DatIAD$Study == "Hav20STZ", ])
summary(NSSpost_4)

```

Comparing untreated MCAO and sham: Significantly lower impairment (NSS) for sham in all studies.  
Comparing untreated MCAO and JM-20: No significant differences but in a single study (Hav19).



### Meta-analysis of post NSS

We now need to gather the results from the separate regression models, ie, their coefficients. We can simply create a new object in R that contains these results.

__*Exercise:*__  
Complete the listings of four numbers each from the previous regressions. Remember, meta-analysis always needs an effect-measure __and__ a measure of uncertainty (called standard error, SE, or variance):

```{r gatherEffectsNaive, exercise=TRUE}
mean_post <- c(1, -7.125, , )
se_mean_post <- c(0.61, 1.74, , )
```

```{r gatherEffectsNaive-solution}
mean_post <- c(1, -7.125, -2.26, 0.18)
se_mean_post <- c(0.61, 1.74, 2.17, 1.02)
```


We could also do this more efficient, where we would need no copy paste:

```{r gather effects, echo=TRUE, message=FALSE, warning=FALSE}

mean_post <- c(summary(NSSpost_1)$coefficients[2,1], # extract specific estimate: mean diff JM20 vs MCAO-control
               summary(NSSpost_2)$coefficients[2,1],
               summary(NSSpost_3)$coefficients[2,1],
               summary(NSSpost_4)$coefficients[2,1])
se_mean_post <- c(summary(NSSpost_1)$coefficients[2,2],
                  summary(NSSpost_2)$coefficients[2,2],
                  summary(NSSpost_3)$coefficients[2,2],
                  summary(NSSpost_4)$coefficients[2,2])
```


__*Exercise:*__  

You are now ready to conduct the second step, the meta-analysis itself. metafor uses a simple command called rma(), which takes the listed results from above. It calls the effects "y" and theLook at the output of this analysis, and forward to a (default) forest plot.

```{r twoStepMA, exercise=TRUE}

mean_post_ma <- rma(yi = ,
                    sei = )
summary()

forest(,
       slab = c("Berlin19", "Hav19", "Hav20AGED", "Hav20STZ"),
       xlab = "mean difference NSS after 7d")

```

Aggregated effect mean difference in NSS at day 7: -1.80, ie. slightly lower NSS in JM-20 group than in untreated MCAO control. Non-significant.
